\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-11-01 demystify.tex
updates

\section{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply?
\begin{itemize}
    \item Yes, definitely.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools.
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by a committee and each application is reviewed by typically 3 faculty.
    \item Individually faculty cannot directly admit a student (so do not email and ask if you have a chance), but they can \emph{advocate} for a student (so do email and introduce yourself if you share the same research interest).
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability.
    \item GRE \emph{is not} required by most schools.
    \item Grades are important, but only if your school is well-known. Otherwise, they are not useful.
\end{itemize}
\end{enumerate}

\begin{itemize}
    \item 
\end{itemize}

\section{Should You Apply?}
\epigraph{\vspace{-0.2in} Don't make fun of graduate students. They just made a terrible life choice.}{\textsc{The Simpsons}}

@@

\subsection{Is an MS degree required for admission to PhD in CS?}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

@@

\subsection{How long does it take to complete the CS PhD program?}

@@

    The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th year where the ambition is \emph{``Just let me graduate''}).

@@

@@

>>>>

\section{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a PhD in CS in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\end{enumerate}

\section{Should You Apply to do a PhD in CS in the US?}\label{sec:should}

\epigraph{\vspace{-0.2in} Don't make fun of graduate students. They just made a terrible life choice.}{\textsc{The Simpsons}}

@@

\subsection{Is an MS degree required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

@@

\subsection{How long does it take to complete the CS PhD program?}\label{sec:time}

@@

    The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year where the ambition is \emph{``Just let me graduate''}).

\subsection{Difference between PhD in the US and other countries}
Here I try to list the main differences between CS PhD in the US and other countries.  This is based on my experience in the US and what I heard from others who did PhD in other countries.

\emph{MS Degree requirement}:  as mentioned in \S\ref{sec:msrequirement} and \ref{sec:time}, US PhD programs do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

\emph{Project proposal}: in many countries, you have to choose a project and advisor \emph{during} the application process (e.g., you write a proposal for potential advisor). In the US, you start your PhD without a project and advisor and find them later. This is also why US PhD programs are longer.

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential advisor and research topics. In other countries, you often do not do course work and start your research right away, e.g., you immediately work on the research project you propose with the advisor you chose. In the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense.  In other countries, you do not have to do these exams or only do few of them. For example, ABD (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.

\emph{Funding}:  In Europe, stipend (salary) comes from the university or from Gov research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your advisor (and no fixed duration).  There are fewer TA opportunities in the European universities compared to the US.

\emph{Faculty Position after PhD}: In other countries, PhD graduates typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates interested in academia can apply directly for faculty position.  This is why US PhD programs are longer and more focused on research.

\emph{Work-life balance}: PhD students are often overworked in the US (seriously!) compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to work for the dept. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA work.

@@

      \begin{center}
        \includegraphics[scale=0.5]{c6.png}
      \end{center}

@@

\subsection{Will I be miserable during my a PhD?}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable.  Sure there are issues with PhD and academic in general, but then there are also issues any jobs, family, and life.  Certainly there are horrible professors, bad relationships with co-authors and officemates, feeling left out or discriminated (e.g., because you're an international student), and it is good to be aware of those.  However, in reality there are many good mentors, supportive labmates and department, happy students, etc.  Don't let social media make you feel pessimistic and destroy your quest to advance knowledge.
